Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-19 paper/paper.tex
Cite tapl; fix universe hierarchy

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{A Short and Mechanized Consistency Proof for Dependent Type Theory}

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize stronger
results\scw{stronger than what?} related to the decidability of type conversion or type
checking of dependently typed systems, from which consistency can be
derived as a corollary.

@@

Introductory materials about logical relations such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical}
start from logical relations for simply typed languages,\scw{proving 
results such as \ldots} and
eventually extend the technique to build a relational model for System
F in order to derive
parametricity.
\scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
Does Aspinall and Hofmann's chapter of ATTAPL include a logical relation?}
\scw{Seems like you should also mention POPLmark reloaded somehere too?}
 \citet{harperkripke} gives a gentle introduction on how
logical relations for closed terms can be extended to talk about open
terms so we can derive properties such as normalization for well-typed
open terms. Overall, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

@@

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
\scw{Is the precondition for T-ABS really admissible? Isn't it required 
to ensure predicativity? Have you proved it? Have you proved regularity?}
\scw{Do you have a formation rule for identity types?}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Short and Mechanized Logical Relation for Dependent Type Theory}

@@

The consistency proofs of various dependently typed systems, including Martin-Lof's
type theory and the Calculus of Constructions, have long been
available in the literature. In particular, recent works such as \citet{nbeincoq},
\citet{decagda}, and \citet{martin-lof-a-la-coq} mechanize the
correctness of the NbE algorithm, the decidability of type
conversion, decidability of type
checking respectively for dependently typed systems. From these
properties, consistency can be derived as a corollary.

@@

Introductory materials about logical relations or standard textbooks such as
\citet{skorstengaard2019introduction}, \citet{harper2016practical},
and \citet{pierce2002types}
talk about logical relations for simply or polymorphically typed
languages, proving results such as termination for closed terms and parametricity.
\scw{Are there some OPLSS notes from Amal Ahmed to also cite? Other modern textbooks?
Does Aspinall and Hofmann's chapter of ATTAPL include a logical
relation?}
\yl{skorstengaard is from the OPLSS notes. Cited ATTAPL}
\yl{Don't have a copy of }
\scw{Seems like you should also mention POPLmark reloaded somehere too?}
\citet{pierce2004advanced}, \citet{harperkripke}, and \citet{abel2019poplmark} show how
a Kripke-style logical relation can be used to include scoping
information in the logical relation and
derive properties such as the existence of normal form or strong
normalization for open and closed terms in simply typed languages.
Overall, the introductory texts about logical relations
cover systems and properties with varying degrees of complexity, from
simply typed to polymorphicly typed, logical predicate to logical
equivalence, closed terms to open terms.

@@

\begin{figure}[ht]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
\scw{Is the precondition for T-ABS really admissible? Isn't it required 
to ensure predicativity? Have you proved it? Have you proved regularity?}
\yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
  was formulated correctly (which wasn't the case in the version
  of the paper you read). I revised \rref{T-Univ} so the premise is
  not admissible }
\scw{Do you have a formation rule for identity types?}
\yl{Added}
